---
title: "Prune the IE and ST phylogenies with PHOIBLE data"
author: "Steven Moran\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
bibliography: '../bibliography/references.bib'  
---

# Overview

We use [R](https://www.r-project.org/) [@R] and the following [R packages](https://cran.r-project.org/web/packages/available_packages_by_name.html) [@testthat;@phylotools] in this script:

```{r message=FALSE}
library(testthat)
library(phylotools)
source('lib/functions.R')

# Choose which traits to use to prune:
# load("trees/traits.Rdata")
load("../Harald/traits-harald.Rdata")
```

We prune the language phylogeneies in our sample with the daughter languages that we have in the traits data that we identified in PHOIBLE [@MoranMcCloy2019]. This only needs to be run once (per input source, i.e., phoible traits and extended phoible traits via data mining from Harald Hammarström; see main text in the paper for details).


## Indo-European

The Indo-European phylogeny was published by @Changetal2015Ancestry-constrained and is available in [D-PLACE](https://github.com/D-PLACE/dplace-data) [@kirby2016d].
  
  
```{r}
# Tree paths
tree <- 'trees/ie-c-tree.nex'
forest <- 'trees/ie-c-trees.nex'

# Prune the trees
pr_sum_tree <- PruneSummaryTree(tree, traits$taxa, 'Glottocode')
pr_trees <- PruneForest(forest, traits$taxa, 'Glottocode')

# Check that the summary tree tip labels are the same as (the first) posterior tree tip labels
expect_true(setequal(pr_sum_tree$tip.label, pr_trees$STATE_0$tip.label))

# Prune the traits data to match the tree tips for analysis
data <- PruneTraits(traits, pr_sum_tree$tip.label)

# Combine them into a list of R data objects for analysis with the BT3 wrapper
pr_sum_tree <- list(data=data, tree=pr_sum_tree)
pr_trees <- list(data=data, tree=pr_trees)
# save(pr_sum_tree, pr_trees, file='trees/ie-c-trees.Rdata')
save(pr_sum_tree, pr_trees, file='trees/harald-ie-c-trees.Rdata')
```


## Sino-Tibetan

The Sino-Tibetan phylogeny was published by @Zhang2019 and is available in [D-PLACE](https://github.com/D-PLACE/dplace-data) [@kirby2016d].


```{r}
# Tree paths
tree <- 'trees/sinotibetan-z-tree.nex'
forest <- 'trees/sinotibetan-z-trees.nex'

# Prune the trees
pr_sum_tree <- PruneSummaryTree(tree, traits$taxa, 'Glottocode')
pr_trees <- PruneForest(forest, traits$taxa, 'Glottocode')

# Check that the summary tree tip labels are the same as (the first) posterior tree tip labels
expect_true(setequal(pr_sum_tree$tip.label, pr_trees$STATE_0$tip.label))

# Prune the traits data to match the tree tips for analysis
data <- PruneTraits(traits, pr_sum_tree$tip.label)

# Combine them into a list of R data objects for analysis with the BT3 wrapper
pr_sum_tree <- list(data=data, tree=pr_sum_tree)
pr_trees <- list(data=data, tree=pr_trees)
# save(pr_sum_tree, pr_trees, file='trees/sinotibetan-z-trees.Rdata')
save(pr_sum_tree, pr_trees, file='trees/harald-sinotibetan-z-trees.Rdata')
```

Clean up.

```{r}
rm(pr_sum_tree, pr_trees, tree, forest)
```


# References

