---
title: "Phylogenetic study of ejectives and uvulars (presense / absence) in IE and ST"
author: "Steven Moran\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
bibliography: '../bibliography/references.bib'
---

# Overview

For the traits, we use presence vs. absence of ejectives and uvulars. Hence, we can use the "Nonmarginal Ejective" and "Nonmarginal Uvular" columns in the datasheet. These are the number of ejectives/uvulars minus number of marginal ejectives/uvulars. For the logistic regression, these counts were used as the basis for the binary distinction. So if there is one uvular but it is coded as marginal then it is going to be noted as FALSE.

For the phylognetic traits, we use presence / absence of class, i.e., a language does or does not have ejectives or uvulars. This is basically a discrete variable model for phylogenetic reconstruction. Then we’ll have the probability of the trait (variable) at the root. We use the Indo-European and Sino-Tibetan phylogenies published by [@Changetal2015Ancestry-constrained] and [@Zhang2019], respectively, and available in [D-PLACE](https://github.com/D-PLACE/dplace-data) [@kirby2016d].

We use [R](https://www.r-project.org/) [@R] and the following [R packages](https://cran.r-project.org/web/packages/available_packages_by_name.html) [@dplyr;@knitr;@testthat]:

```{r message=FALSE}
library(dplyr)
library(knitr)
library(testthat)
```


# Data prep

Data inspection and cleaning.

```{r}
df <- read.csv('../Data/uvulars_ejectives_pruned2.csv', stringsAsFactors = FALSE)
```

Which sources are <NA> for Glottocodes? This issue is discussed in detail here: 

* https://github.com/phoible/dev/issues/270

Drop NAs.

```{r}
df %>% filter(is.na(Glottocode))
df <- df %>% filter(!is.na(Glottocode))
expect_equal(nrow(df %>% filter(is.na(Glottocode))), 0)
```

Extract relevant data for this analysis and then investigate ejectives and uvulars.

```{r}
ejectives <- df %>% select(InventoryID, Source, Glottocode, Nonmarginal_Ejective) %>% rename(has_ejectives = Nonmarginal_Ejective)
uvulars <- df %>% select(InventoryID, Source, Glottocode, Nonmarginal_Uvular) %>% rename(has_uvulars = Nonmarginal_Uvular)
```


## Ejectives

Check for duplicate Glottocodes (we can have only one data point per daughter node on the phylogeny).

```{r}
e.dups <- ejectives %>% group_by(Source, Glottocode) %>% filter(n()>1)
e.dups %>% arrange(Glottocode) %>% kable()
```

Drop some of these data points.

```{r}
ejectives <- ejectives %>% filter(!(InventoryID %in% c(2883, 2037, 2031, 2207, 2345,2348, 2618, 2988)))
expect_equal(nrow(ejectives %>% group_by(Source, Glottocode) %>% filter(n()>1)), 0)
```


## Uvulars

Check for duplicate Glottocodes (we can have only one data point per daughter node on the phylogeny).

```{r}
u.dups <- uvulars %>% group_by(Source, Glottocode) %>% filter(n()>1)
u.dups %>% arrange(Glottocode) %>% kable()
```

There's disagreement about the presence of absence of uvulars between inventories for:

```{r}
u.dups %>% filter(Glottocode %in% c('ajab1235', 'basq1248', 'mose1249', 'nort2980', 'port1283', 'stan1288', 'west2368')) %>% arrange(Glottocode) %>% kable() 
```

One inventory for ajab1235 lists uvulars the second does not. This may because these are different dialects with the same Glottocode. It is clear from the first inventory, and a subsequent publication by Eric Morey ("A Grammar of Ajagbe") that it contains two uvulars. We take thus the first inventory.

* https://phoible.org/inventories/view/651
* https://phoible.org/inventories/view/652

SPA inventory for basq1248 lists a uvular, the other does not. SPA lists it as a loan, so we take the second inventory.

* https://phoible.org/inventories/view/179
* https://phoible.org/inventories/view/2161

The inventories for Moseten are:

* https://phoible.org/inventories/view/944
* https://phoible.org/inventories/view/1986

The first contains uvular and Sakel (2004:30) states:

```
/h/:[X] and [h]

The velar fricative [X] and the uvular fricative [h] form one phoneme /h/. [h] exclusively occurs at the beginning of a syllable.
```

The IPA chart (page 25) lists glottal fricative /h/ with a footnote (36: Allophone: [x]). I think should be updated to be in line with the SAPHON interpretation.

The inventories for Quechua include:

* https://phoible.org/inventories/view/2030
* https://phoible.org/inventories/view/2037
* https://phoible.org/inventories/view/2031

The first two indicate one uvular.

Portuguese is listed in PHOIBLE twice:

* https://phoible.org/inventories/view/2206
* https://phoible.org/inventories/view/2207

The European dialect is noted as having a uvular. As per Wikipedia:

* https://en.wikipedia.org/wiki/Portuguese_phonology

```
The consonant hereafter denoted as /ʁ/ has a variety of realizations depending on dialect. In Europe, it is typically a uvular trill [ʀ]; however, a pronunciation as a voiced uvular fricative [ʁ] may be becoming dominant in urban areas. There is also a realization as a voiceless uvular fricative [χ], and the original pronunciation as an alveolar trill [r] also remains very common in various dialects.[15] A common realization of the word-initial /r/ in the Lisbon accent is a voiced uvular fricative trill [ʀ̝].[16] In Brazil, /ʁ/ can be velar, uvular, or glottal and may be voiceless unless between voiced sounds;[17] it is usually pronounced as a voiceless velar fricative [x], a voiceless glottal fricative [h] or voiceless uvular fricative [χ]. See also Guttural R in Portuguese.
```

Spanish also has two inventories:

* https://phoible.org/inventories/view/2308
* https://phoible.org/inventories/view/2210

One with uvular (the Castilian variety).

Lastly, different dialects of Western Balochi:

* https://phoible.org/inventories/view/2622
* https://phoible.org/inventories/view/2345
* https://phoible.org/inventories/view/2348
* https://phoible.org/inventories/view/2618

the first reports a uvular, the others do not.

For the time being, we drop the same inventories as we did with the ejectives.

```{r}
uvulars <- uvulars %>% filter(!(InventoryID %in% c(2883, 2037, 2031, 2207, 2345, 2348, 2618, 2988)))
expect_equal(nrow(uvulars %>% group_by(Source, Glottocode) %>% filter(n()>1)), 0)
```

Note that we still have duplicate Glottocodes across sources.

```{r}
ejectives %>% group_by(Glottocode) %>% filter(n()>1) %>% kable()
```

```{r}
uvulars %>% group_by(Glottocode) %>% filter(n()>1) %>% kable()
```

We remove the UPSID inventory for the time being given the trump hierarchy.

```{r}
ejectives <- ejectives %>% filter(!(InventoryID %in% c(648)))
uvulars <- uvulars %>% filter(!(InventoryID %in% c(648)))
```

Since we've removed the same data points in the ejectives and uvulars data, make sure that they match.

```{r}
expect_true(all(ejectives$InventoryID %in% uvulars$InventoryID))
expect_true(all(uvulars$InventoryID %in% ejectives$InventoryID))
```

Some clean up.

```{r}
rm(e.dups, u.dups)
```


## Formulate the traits files

First select just the columns for Glottocodes and presence/absence of the variable. Then reformulate as "Y" (present) vs "N" (absent). Lastly, turn the dataframe into something that works in a discrete variable phylogenetic analysis.

```{r}
ejectives.traits <- ejectives %>% select(Glottocode, has_ejectives)
table(ejectives.traits$has_ejectives)
ejectives.traits <- mutate_at(ejectives.traits, vars(-Glottocode), function(x) ifelse(x==0, "N", "Y"))
rownames(ejectives.traits) <- ejectives.traits[,1]
colnames(ejectives.traits)[1] <- "taxa"
head(ejectives.traits)
table(ejectives.traits$has_ejectives)
```

```{r}
uvulars.traits <- uvulars %>% select(Glottocode, has_uvulars)
table(uvulars.traits$has_uvulars)
uvulars.traits <- mutate_at(uvulars.traits, vars(-Glottocode), function(x) ifelse(x==0, "N", "Y"))
rownames(uvulars.traits) <- uvulars.traits[,1]
colnames(uvulars.traits)[1] <- "taxa"
head(uvulars.traits)
table(uvulars.traits$has_uvulars)
```

Combine the traits files so that the languages can be analyzed the distribution of ejectives and uvulars.

```{r}
traits <- left_join(ejectives.traits, uvulars.traits)
rownames(traits) <- traits[,1]
```

Here is the cross-tabular distribution of ejectives vs uvulars.

```{r}
table(traits$has_ejectives, traits$has_uvulars) %>% kable()
```

Save the traits dataframes.

```{r}
save(traits, ejectives.traits, uvulars.traits, file="trees/traits.Rdata")
```


# References
