---
title: "Phylogenetic study of ejectives and uvulars (presense / absence) in IE and ST"
author: "Steven Moran"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
---
  
# Analyses

For each prune phylogeny we plot the trait coverage (has or does not have ejectives and uvulars) and then we generate stochasic character maps of each trait for each phylogeny.

Each phylogeny has been pruned for the data points that it contains in PHOIBLE 2.0 (Moran & McCloy, 2019). The traits (presence or absence of ejectives or uvulars) in are then plotted as present (green) or not present (red) in the languages represented in the phylogeny by Glottocodes (Hammarström et al, 2020).

References

* Moran, Steven & McCloy, Daniel (eds.) 2019.PHOIBLE 2.0. Jena: Max Planck Institute for the Science of Human History. (Available online at http://phoible.org, Accessed on 2020-05-16.)

* Hammarström, Harald & Forkel, Robert & Haspelmath, Martin & Bank, Sebastian. 2020. Glottolog 4.2.1. Jena: Max Planck Institute for the Science of Human History. (Available online at http://glottolog.org, Accessed on 2020-05-16.)

To generate stochastic character maps (Nielsen, 2002; Huelsenbeck et. al, 2003; Revell, 20120) we use the `phytools::make.simmap` function for an all rates different (ARD) model where `q` is set to `empirical` (maximum probability, full Bayesian MCMC) with 10 simulations on the ACQDIV phonology data.

References

* R. Nielsen, Mapping mutations on phylogenies. Syst. Biol. 51, 729–739 (2002). doi: 10.1080/10635150290102393;
pmid: 12396587

* J. P. Huelsenbeck, R. Nielsen, J. P. Bollback, Stochastic mapping of morphological characters. Syst. Biol. 52, 131–158 (2003). doi: 10.1080/10635150390192780; pmid: 12746144

* L. J. Revell, phytools: An R package for phylogenetic comparative biology (and other things). Methods Ecol. Evol. 3, 217–223 (2012). doi: 10.1111/j.2041-210X.2011.00169.x


```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggtree)
library(ggplot2) # http://bioconductor.org/packages/release/bioc/html/ggtree.html

library(phytools)
library(phylotools)
source('lib/functions.R')

load("../Harald/traits-harald.Rdata")
```

Plot function.

```{r}
# Define color schema
color.scheme <- c('green','red')
names(color.scheme) <- c('Y','N')

# Function to reverse time in the plot
reverse.time <- function(p) { p$data$x <- p$data$x - max(p$data$x); return(p) }

# Create tree and heatmap figure
plot.tree <- function(pr_sum_tree_plot, features_plot) {
  gheatmap(pr_sum_tree_plot, features_plot,
           colnames_position='top', width=2.2, offset=2000, color='black',
           colnames_offset_y = 0.4, font.size = 2.5, colnames_angle = 40) +
    scale_fill_manual(name="", values=color.scheme) + 
    scale_x_continuous(breaks=c(-6000,-4000,-2000,0)) +
    scale_y_continuous(expand = c(-0.01, 1)) +
    theme_tree2(axis.text.x=element_text(size=8)) +
    theme(legend.position='none', axis.ticks.length=unit(4,"mm"), 
          axis.ticks=element_line(color='grey'))
}
```


## Indo-European

The phylogeny comes from:
  
  * Chang W, Cathcart C, Hall D, & Garrett A. 2015. Ancestry-constrained phylogenetic analysis supports the Indo-European steppe hypothesis. Language, 91(1):194-244.

Here are the traits plotted on the pruned phylogeny. 

```{r message=FALSE, warning=FALSE}
load('trees/harald-ie-c-trees.Rdata')
traits.print <- pr_sum_tree$data %>% select(-taxa)
p <- reverse.time(ggtree(pr_sum_tree$tree, ladderize=T, right=T)) +
  geom_tiplab(align=T, linesize = .1, size = 1.5) 
plot.tree(p, traits.print)
```

With this data, the IE pruned tree has this many data points:

```{r}
nrow(pr_sum_tree$data)
```


Here are the stochastic character maps.

For ejectives:

```{r}
data <- pr_sum_tree$tree[["tip.label"]]
data <- data.matrix(data)

data <- pr_sum_tree$data[, "has_ejectives"]
names(data) <- pr_sum_tree$data$taxa

# Generate simmmap data: empirical = q has the maximum probablity; full bayesian = mcmc, pi=estimated
treemaps <- make.simmap(pr_sum_tree$tree, data, model="ARD", nsim=10, Q="empirical", pi="estimated")

# Plot simmap
map = densityMap(treemaps, plot=F, res=300)
map$cols[1:length(map$cols)] <- rev(map$cols[1:length(map$cols)])
p <- plot(map, outline=T, lwd=c(3,3),
          fsize=c(0.6,0.6), ftype='reg', 
          legend=2000, leg.txt=c("0",paste0("Posterior Probability (Y)"),"1"))
title(main="Ejectives")
p
```

For uvulars:

```{r}
data <- pr_sum_tree$tree[["tip.label"]]
data <- data.matrix(data)

data <- pr_sum_tree$data[, "has_uvulars"]
names(data) <- pr_sum_tree$data$taxa

# Generate simmmap data: empirical = q has the maximum probablity; full bayesian = mcmc, pi=estimated
treemaps <- make.simmap(pr_sum_tree$tree, data, model="ARD", nsim=10, Q="empirical", pi="estimated")

# Plot simmap
map = densityMap(treemaps, plot=F, res=300)
map$cols[1:length(map$cols)] <- rev(map$cols[1:length(map$cols)])
p <- plot(map, outline=T, lwd=c(3,3),
          fsize=c(0.6,0.6), ftype='reg', 
          legend=2000, leg.txt=c("0",paste0("Posterior Probability (Y)"),"1"))
title(main="Uvulars")
p
```


## Sino-Tibetan

Phylogeny comes from:
  
  * Zhang M, Yan S, Pan W, & Jin L. 2019. Phylogenetic evidence for Sino-Tibetan origin in northern China in the Late Neolithic. Nature, 569, 112-115.

Here are the traits plotted on the pruned phylogeny.

```{r message=FALSE, warning=FALSE}
load('trees/harald-sinotibetan-z-trees.Rdata')
traits.print <- pr_sum_tree$data %>% select(-taxa)
p <- reverse.time(ggtree(pr_sum_tree$tree, ladderize=T, right=T)) +
  geom_tiplab(align=T, linesize = .1, size = 1.5) 
plot.tree(p, traits.print)
```

With this data, the ST pruned tree has this many data points:

```{r}
nrow(pr_sum_tree$data)
```


```{r}
data <- pr_sum_tree$tree[["tip.label"]]
data <- data.matrix(data)

data <- pr_sum_tree$data[, "has_ejectives"]
names(data) <- pr_sum_tree$data$taxa

# Generate simmmap data: empirical = q has the maximum probablity; full bayesian = mcmc, pi=estimated
treemaps <- make.simmap(pr_sum_tree$tree, data, model="ARD", nsim=10, Q="empirical", pi="estimated")

# Plot simmap
map = densityMap(treemaps, plot=F, res=300)
map$cols[1:length(map$cols)] <- rev(map$cols[1:length(map$cols)])
p <- plot(map, outline=T, lwd=c(3,3),
          fsize=c(0.6,0.6), ftype='reg', 
          legend=2000, leg.txt=c("0",paste0("Posterior Probability (Y)"),"1"))
title(main="Ejectives")
p
```

For uvulars:

```{r}
data <- pr_sum_tree$tree[["tip.label"]]
data <- data.matrix(data)

data <- pr_sum_tree$data[, "has_uvulars"]
names(data) <- pr_sum_tree$data$taxa

# Generate simmmap data: empirical = q has the maximum probablity; full bayesian = mcmc, pi=estimated
treemaps <- make.simmap(pr_sum_tree$tree, data, model="ARD", nsim=10, Q="empirical", pi="estimated")

# Plot simmap
map = densityMap(treemaps, plot=F, res=300)
map$cols[1:length(map$cols)] <- rev(map$cols[1:length(map$cols)])
p <- plot(map, outline=T, lwd=c(3,3),
          fsize=c(0.6,0.6), ftype='reg', 
          legend=2000, leg.txt=c("0",paste0("Posterior Probability (Y)"),"1"))
title(main="Uvulars")
p
```



