---
title: Bayesian modelling of ejectives and uvulars depending on altitude and ancillary
  analyses
author: "Matthias Urban"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document: default
---
# Overview

This document describes the Bayesian modeling of ejectives and uvulars (presence or absence) depending on altitude and ancillary analyses

# Required packages

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(boot)
```

# Data preparation

Read in data

```{r}
elevdata <- read.csv("../Data/uvulars_ejectives_pruned2_rhotics.csv", header = T)
```

Change separator from comma to dot for coordinates and treat as numeric

```{r}
elevdata$latitude <- as.numeric(gsub(",", ".", elevdata$latitude))
elevdata$longitude <- as.numeric(gsub(",", ".", elevdata$longitude))
```

Treat macroareas as factors

```{r}
elevdata$macroarea2 <- as.factor(elevdata$macroarea2)
```

Remove rows with empty cells for Latitude and Longitude

```{r}
elevdata <- drop_na(elevdata, elevation)
```

Reduce the data to a binary distinction between presence vs. absence of ejectives/uvulars

```{r}
elevdata <- mutate(elevdata, NonMarginal01 = as.logical(Nonmarginal_Uvular), NonMarginal01 = as.numeric(NonMarginal01))
elevdata <- mutate(elevdata, NonMarginal02 = as.logical(Nonmarginal_Ejective), NonMarginal02 = as.numeric(NonMarginal02))
elevdata <- mutate(elevdata, NonMarginal03 = as.logical(Nonmarginal_Uvular_no_rhotics), NonMarginal03 = as.numeric(NonMarginal03))
```

# Modeling


## Bayesian logistic mixed effects regressions

Tranform elevation to its log10 for modelling
```{r, warning=FALSE}
elevdata <- mutate(elevdata, elevationlog10 = log10(elevation))
```

Set prior
```{r}
priors <- set_prior("normal(0, 2)", class = "b", coef = "elevationlog10")
```

### Model for uvulars
```{r, warning=FALSE, results='hide', cache=TRUE}
elevmodeluvulars <- brm(NonMarginal01 ~ elevationlog10 + (1 + elevationlog10 | macroarea2) + (1 | family_id), family = "bernoulli", data = elevdata, seed = 31011, warmup = 6000, iter = 8000, chains = 4, prior = priors, control = list(adapt_delta = 0.999, max_treedepth = 20))
```

Model assessment and checks

check Rhat and ESS values
```{r}
summary(elevmodeluvulars)
```
Inspect chains
```{r}
plot(elevmodeluvulars)
```

Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(elevmodeluvulars)
pp_check(elevmodeluvulars, type = "error_binned")
```

Assess predictive accuracy
```{r}
modelled_elevdata <- elevdata %>% drop_na(NonMarginal01, elevationlog10, macroarea2, family_id)
elevmodeluvulars_pred <- predict(elevmodeluvulars, type = "response")[, "Estimate"]
elevmodeluvulars_pred <- as.numeric(elevmodeluvulars_pred > mean(modelled_elevdata$NonMarginal01))
(classtab_elevmodeluvulars <- table(predicted = elevmodeluvulars_pred, observed = modelled_elevdata$NonMarginal01))
(acc_elevmodeluvulars <- sum(diag(classtab_elevmodeluvulars)) / sum(classtab_elevmodeluvulars))
```

Assess posterior probability versus chance
```{r}
elevmodeluvularssamples <- posterior_samples(elevmodeluvulars)
sum(elevmodeluvularssamples$b_elevationlog10 < 0) / nrow(elevmodeluvularssamples)
```

### Model for uvulars without rhotics
```{r, warning=FALSE, results='hide', cache=TRUE}
elevmodeluvularswithoutrhotics <- brm(NonMarginal03 ~ elevationlog10 + (1 + elevationlog10 | macroarea2) + (1 | family_id), family = "bernoulli", data = elevdata, seed = 31011, warmup = 6000, iter = 8000, chains = 4, prior = priors, control = list(adapt_delta = 0.999, max_treedepth = 20))
```

Model assessment and checks

check Rhat and ESS values

```{r}
summary(elevmodeluvularswithoutrhotics)
```

Inspect chains
```{r}
plot(elevmodeluvularswithoutrhotics)
```

Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(elevmodeluvularswithoutrhotics)
pp_check(elevmodeluvularswithoutrhotics, type = "error_binned")
```

Assess predictive accuracy
```{r}
modelled_elevdata <- elevdata %>% drop_na(NonMarginal03, elevationlog10, macroarea2, family_id)
elevmodeluvularswithoutrhotics_pred <- predict(elevmodeluvularswithoutrhotics, type = "response")[, "Estimate"]
elevmodeluvularswithoutrhotics_pred <- as.numeric(elevmodeluvularswithoutrhotics_pred > mean(modelled_elevdata$NonMarginal03))
(classtab_elevmodeluvularswithoutrhotics <- table(predicted = elevmodeluvularswithoutrhotics_pred, observed = modelled_elevdata$NonMarginal03))
(acc_elevmodeluvularswithoutrhotics <- sum(diag(classtab_elevmodeluvularswithoutrhotics)) / sum(classtab_elevmodeluvularswithoutrhotics))
```

Assess posterior probability versus chance
```{r}
elevmodeluvularswithoutrhoticssamples <- posterior_samples(elevmodeluvularswithoutrhotics)
sum(elevmodeluvularswithoutrhoticssamples$b_elevationlog10 < 0) / nrow(elevmodeluvularswithoutrhoticssamples)
```

### Model for ejectives
```{r, warning=FALSE, results='hide', cache=TRUE}
elevmodelejectives <- brm(NonMarginal02 ~ elevationlog10 + (1 + elevationlog10 | macroarea2) + (1 | family_id), family = "bernoulli", data = elevdata, seed = 31011, warmup = 6000, iter = 8000, chains = 4, prior = priors, control = list(adapt_delta = 0.999, max_treedepth = 20))
```

Model assessment and checks

Check Rhat and ESS values
```{r}
summary(elevmodelejectives)
```

Inspect chains
```{r}
plot(elevmodelejectives)
```

Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(elevmodelejectives)
pp_check(elevmodelejectives, type = "error_binned")
```

Assess predictive accuracy
```{r}
modelled_elevdata <- elevdata %>% drop_na(NonMarginal02, elevationlog10, macroarea2, family_id)
elevmodelejectives_pred <- predict(elevmodelejectives, type = "response")[, "Estimate"]
elevmodelejectives_pred <- as.numeric(elevmodelejectives_pred > mean(modelled_elevdata$NonMarginal02))
(classtab_elevmodelejectives <- table(predicted = elevmodelejectives_pred, observed = modelled_elevdata$NonMarginal02))
(acc_elevmodelejectives <- sum(diag(classtab_elevmodelejectives)) / sum(classtab_elevmodelejectives))
```

Assess posterior probability versus chance
```{r}
elevmodelejectivessamples <- posterior_samples(elevmodelejectives)
sum(elevmodelejectivessamples$b_elevationlog10 < 0) / nrow(elevmodelejectivessamples)
```

## By-area and by-family analysis

### By area
Compute median elevations and proportions of uvulars and ejectives by area
```{r}
medianelevarea <- aggregate(elevation ~ macroarea2, FUN = "median", data = elevdata)
uvularproportionarea <- aggregate(NonMarginal01 ~ macroarea2, FUN = "mean", data = elevdata)
uvularwithoutrhoticsproportionarea <- aggregate(NonMarginal03 ~ macroarea2, FUN = "mean", data = elevdata)
ejectiveproportionarea <- aggregate(NonMarginal02 ~ macroarea2, FUN = "mean", data = elevdata)
```

Least squares regression
```{r}
summary(lm(uvularproportionarea$NonMarginal01 ~ medianelevarea$elevation))
summary(lm(uvularwithoutrhoticsproportionarea$NonMarginal03 ~ medianelevarea$elevation))
summary(lm(ejectiveproportionarea$NonMarginal02 ~ medianelevarea$elevation))
```

Plot results
```{r}
plot(medianelevarea$elevation, uvularproportionarea$NonMarginal01, bg = "black", pch = 21, xlab = "Median elevation per area", ylab = "Proportion Uvulars")
lines(lowess(medianelevarea$elevation, uvularproportionarea$NonMarginal01, f = 10, iter = 10))
plot(medianelevarea$elevation, uvularwithoutrhoticsproportionarea$NonMarginal03, bg = "black", pch = 21, xlab = "Median elevation per area", ylab = "Proportion Uvulars without Rhotics")
lines(lowess(medianelevarea$elevation, uvularwithoutrhoticsproportionarea$NonMarginal03, f = 10, iter = 10))
plot(medianelevarea$elevation, ejectiveproportionarea$NonMarginal02, bg = "black", pch = 21, xlab = "Median elevation per area", ylab = "Proportion Ejectives")
lines(lowess(medianelevarea$elevation, ejectiveproportionarea$NonMarginal02, f = 10, iter = 10))
```

### By family

```{r}
largefamilies <- filter(elevdata, family_id %in% c("afro1255", "araw1281", "atha1245", "atla1278", "aust1307", "cari1283", "gong1255", "mand1469", "maya1287", "mong1329", "nakh1245", "otom1299", "sali1255", "sino1245", "taik1256", "tupi1275", "turk1311", "ural1272"))
```

Compute median elevations and proportions of uvulars and ejectives per by family
```{r}
medianelevfamily <- aggregate(elevation ~ family_id, FUN = "median", data = largefamilies)
uvularproportionfamily <- aggregate(NonMarginal01 ~ family_id, FUN = "mean", data = largefamilies)
uvularwithoutrhoticsproportionfamily <- aggregate(NonMarginal03 ~ family_id, FUN = "mean", data = largefamilies)
ejectiveproportionfamily <- aggregate(NonMarginal02 ~ family_id, FUN = "mean", data = largefamilies)
```

Least squares regression
```{r}
summary(lm(uvularproportionfamily$NonMarginal01 ~ medianelevfamily$elevation))
summary(lm(uvularwithoutrhoticsproportionfamily$NonMarginal03 ~ medianelevfamily$elevation))
summary(lm(ejectiveproportionfamily$NonMarginal02 ~ medianelevfamily$elevation))
```

Plot results
```{r}
plot(medianelevfamily$elevation, uvularproportionfamily$NonMarginal01, bg = "black", pch = 21, xlab = "Median elevation per family", ylab = "Proportion Uvulars")
lines(lowess(medianelevfamily$elevation, uvularproportionfamily$NonMarginal01, f = 10, iter = 10))
plot(medianelevfamily$elevation, uvularwithoutrhoticsproportionfamily$NonMarginal03, bg = "black", pch = 21, xlab = "Median elevation per family", ylab = "Proportion Uvulars without Rhotics")
lines(lowess(medianelevfamily$elevation, uvularwithoutrhoticsproportionfamily$NonMarginal03, f = 10, iter = 10))
plot(medianelevfamily$elevation, ejectiveproportionfamily$NonMarginal02, bg = "black", pch = 21, xlab = "Median elevation per family", ylab = "Proportion Ejectives")
lines(lowess(medianelevfamily$elevation, ejectiveproportionfamily$NonMarginal02, f = 10, iter = 10))
```
