---
title: "PHOIBLE uvular and ejective consonants"
author: "Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output: github_document
bibliography: '../bibliography/references.bib'
---

# Overview

This report extracts [uvulars](https://en.wikipedia.org/wiki/Uvular_consonant) and [ejectives](https://en.wikipedia.org/wiki/Ejective_consonant) from [PHOIBLE](https://phoible.org/) [@MoranMcCloy2019].

The output data format for uvulars and ejectives data is a [CSV](https://en.wikipedia.org/wiki/Comma-separated_values) file with the following columns:

```
InventoryID | Glottocode | ... | Uvulars | Ejectives | Marginal_Uvular | ...
1 | aaaa1111 | ... | 2 | 4 | 1 | ...
```

For uvulars, we extract consonants of all manners of articulation (rather than just stops). For ejectives, we extract all places of articulation. 

The final data file is available in the [uvulars_ejectives_pruned2.csv](../Data/uvulars_ejectives_pruned2.csv) file. Here we also add Urban's extended macroareas.


## Data preparation

This report uses these [R](https://cran.r-project.org/) libraries [@tidyverse,@knitr]:

```{r, message=F, warning=F}
library(tidyverse)
library(knitr)
```

Get the PHOIBLE data and merge in the metadata from [Glottolog](https://glottolog.org/) [@Hammarstrom_etal2020].

PHOIBLE data is released periodically be version because the editors add new data, make reported corrections, etc. Here we use the PHOIBLE dev version from [May 2020](https://github.com/phoible/dev/tree/646f5e4f64bfefb7868bf4a3b65bcd1da243976a).

```{r}
phoible <- read_csv(url("https://github.com/phoible/dev/blob/646f5e4f64bfefb7868bf4a3b65bcd1da243976a/data/phoible.csv?raw=true"), 
                    col_types = c(InventoryID = "i", Marginal = "l", .default = "c"))
```

Merge in [Glottolog 4.1](https://glottolog.org/meta/downloads) data.

```{r}
languoids <- read.csv("glottolog_languoid.csv/languoid.csv", stringsAsFactors = FALSE)
geo <- read.csv("languages_and_dialects_geo.csv", stringsAsFactors = FALSE)
phoible <- left_join(phoible, languoids, by = c("Glottocode" = "id"))
phoible <- left_join(phoible, geo)
rm(geo, languoids)
```

Create a PHOIBLE index.

```{r}
index <- phoible %>%
  select(InventoryID, Glottocode, ISO6393, name, LanguageName, SpecificDialect, Source, family_id, level, status, latitude, longitude, country_ids, macroarea) %>%
  distinct()

index <- index %>% rename(GlottologName = name, PhoibleName = LanguageName)
```

Get uvular consonant counts and their number of marginals.

```{r}
uvulars <- phoible %>% filter(grepl("q|ɢ|ɴ|ʀ|χ|ʁ|ʛ|ʟ̠", Phoneme))
uvular_counts <- uvulars %>%
  group_by(InventoryID) %>%
  summarize(Uvulars = n())
uvular_marginals <- uvulars %>%
  filter(Marginal) %>%
  group_by(InventoryID) %>%
  summarize(Marginal_Uvular = n())
```

Let's write the uvular data to disk, so that we can use it to [remove the rhotics](remove_uvular_rhotics.Rmd), as suggested by a reviewer.

```{r}
write_csv(uvulars, 'uvulars.csv')
```

We note here that these are the languages that contain uvular rhotics. There are 22 occurrences out of 1255 observations.

```{r}
uvulars %>% filter(grepl('ʀ', Phoneme)) %>% select(InventoryID, Glottocode, LanguageName, SpecificDialect, Phoneme, Allophones, Marginal) %>% kable()
```

We also note that the PHOIBLE features +dorsal and +back do not capture uvulars that are palatalized (negates [back]), hence the use of regular expression search above.

```{r}
t1 <- uvulars %>%
  select(Phoneme) %>%
  distinct()
"χʲ" %in% t1$Phoneme
t2 <- phoible %>%
  filter(back == "+" & dorsal == "+") %>%
  select(Phoneme) %>%
  distinct()
"χʲ" %in% t2$Phoneme
setdiff(t1$Phoneme, t2$Phoneme)
rm(t1, t2)
```

Get ejectives counts and their number of marginals.

```{r}
ejectives <- phoible %>% filter(grepl("ʼ", Phoneme))
ejective_counts <- ejectives %>%
  group_by(InventoryID) %>%
  summarize(Ejectives = n())
ejective_marginals <- ejectives %>%
  filter(Marginal) %>%
  group_by(InventoryID) %>%
  summarize(Marginal_Ejective = n())
```

Join the data frames.

```{r}
df <- left_join(index, uvular_counts)
df <- left_join(df, uvular_marginals)
df <- left_join(df, ejective_counts)
df <- left_join(df, ejective_marginals)
rm(ejective_counts, ejective_marginals, uvular_counts, uvular_marginals)
```

Write the results to disk.

```{r}
write_csv(df, "uvulars_ejectives.csv")
```

## Exploratory

Have a look at the data.

```{r}
df %>%
  head() %>%
  kable()
```

Do any languages contain uvulars that are always marginal? Yes.

```{r}
kable(df[which(df$Uvulars == df$Marginal_Uvular), ])
kable(phoible %>% select(InventoryID, Glottocode, LanguageName, Phoneme) %>% filter(InventoryID == 354) %>% filter(grepl("q|ɢ|ɴ|ʀ|χ|ʁ|ʛ|ʟ̠", Phoneme)))
```

Do any languages contain ejectives that are always marginal? Yes.

```{r}
kable(df[which(df$Ejectives == df$Marginal_Ejective), ])
kable(phoible %>% select(InventoryID, Glottocode, LanguageName, Phoneme) %>% filter(InventoryID == 1276) %>% filter(grepl("ʼ", Phoneme)))
```

How many Glottocodes are there in phoible?

```{r}
nrow(phoible %>% select(Glottocode) %>% distinct())
```

How many phoible inventories have uvular consonants?

```{r}
nrow(uvulars %>% select(InventoryID) %>% distinct())
```

How many are marginal?

```{r}
nrow(uvulars %>% filter(Marginal) %>% group_by(InventoryID, Marginal)) # 21 rows
kable(uvulars %>% filter(Marginal) %>% group_by(InventoryID, Marginal))
```

How many phoible inventories have ejectives?

```{r}
nrow(ejectives %>% select(InventoryID) %>% distinct())
```

How many are marginal?

```{r}
nrow(ejectives %>% filter(Marginal) %>% group_by(InventoryID, Marginal)) # 23 rows
kable(ejectives %>% filter(Marginal) %>% group_by(InventoryID, Marginal))
```

How are uvulars distributed in phoible (across duplicate languages)?

```{r}
distribution_uvulars <- uvulars %>%
  group_by(Phoneme) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
kable(distribution_uvulars)

# Order the frequency counts and plot the distribution of uvular consonants in the sample
distribution_uvulars$Phoneme <- factor(distribution_uvulars$Phoneme, levels = distribution_uvulars$Phoneme[order(-distribution_uvulars$count)])

# qplot(distribution_uvulars$Phoneme, distribution_uvulars$count)

p <- ggplot(aes(y = count, x = Phoneme), data = distribution_uvulars) +
  geom_bar(stat = "identity", width = 0.3, color = "black") +
  xlab("Segments") +
  ylab("# of languages") +
  theme_minimal() +
  ggtitle("")
p
```

How are uvulars distributed in phoible (across duplicate languages)?

```{r}
distribution_ejectives <- ejectives %>%
  group_by(Phoneme) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
kable(distribution_ejectives)

# Order the frequency counts and plot the distribution of uvular consonants in the sample
distribution_ejectives$Phoneme <- factor(distribution_ejectives$Phoneme, levels = distribution_ejectives$Phoneme[order(-distribution_ejectives$count)])

# qplot(distribution_ejectives$Phoneme, distribution_ejectives$count)

p <- ggplot(aes(y = count, x = Phoneme), data = distribution_ejectives) +
  geom_bar(stat = "identity", width = 0.3, color = "black") +
  xlab("Segments") +
  ylab("# of languages") +
  theme_minimal() +
  ggtitle("")
p
```

Distribution of uvular consonants per inventory (can't use Glottocode because there are multiple doculects).

```{r}
uvulars_counts <- uvulars %>%
  select(InventoryID, Glottocode, Phoneme, macroarea) %>%
  group_by(InventoryID, Glottocode, macroarea) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

qplot(y = uvulars_counts$count)
```

How are they distributed via macroarea?

```{r}
kable(table(uvulars_counts$macroarea))
```

Distribution of ejective per inventory (can't use Glottocode because there are multiple doculects).

```{r}
ejectives_counts <- ejectives %>%
  select(InventoryID, Glottocode, Phoneme, macroarea) %>%
  group_by(InventoryID, Glottocode, macroarea) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

qplot(y = ejectives_counts$count)
```

How are they distributed via macroarea?

```{r}
kable(table(ejectives_counts$macroarea))
```

# References
