---
title: "Remove uvular rhotics from counts"
author: "Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output: github_document
bibliography: '../bibliography/references.bib'
---

```{r, message=F, warning=F}
library(tidyverse)
library(knitr)
library(testthat)
```

# Overview

An anonymous reviewer notes that our decision to include uvular rhotics requires explicit justification. They would like to see if there is any impact on removing rhotics from the uvular counts on our results.

An important question in response is, what does the reviewer mean explicitly by rhotic? Rhotics are cross-linguistically hard to define; see for example the in-depth description of rhotics in [@LadefogedMaddieson1996]. In particular, what phonetic properties should be considered for the "gluttal R" phenomenon?

* https://en.wikipedia.org/wiki/Guttural_R

Ladefoged & Maddieson mention that the prototypical rhotic members are trills made with the tip or blade of the tongue, i.e., alveolar [r] or uvular [ʀ]. However, other continuants made at the uvular place of articulation can also be categorized as rhotics.

As they point out, from an expert phonetician's point of view, it:

>>> "is whether the class membership is based only on synchronic and diachronic relationships between the members of the class, or whether there is indeed a phonetic similarity between all rhotics that has hitherto been missed"

Additionally, they note that similarity may be either acoustic or auditory.

We show in our [extraction code](get_uvulars_ejectives.md) how many general rhotics, i.e. according the [IPA chart](https://www.internationalphoneticassociation.org/content/full-ipa-chart) are present in the data we extracted.

"having thought a bit more, I'd include ʁ and ʀ but exclude χ and 
cite Maddieson & Ladefoged in support."

Let's drop the rhotics, including [ʁ] and [ʀ], but excluding [χ] [@LadefogedMaddieson1996] and re-generate the counts. First load the uvular data.

```{r, message=F, warning=F}
uvulars <- read_csv("uvulars.csv")
```

Next, get the uvulars without the rhotics.

```{r}
uvulars_no_rhotics <- uvulars %>% filter(!grepl("ʀ|ʁ", Phoneme))
```

Which rhotics did we remove?

```{r}
uvulars_with_rhotics <- uvulars %>% filter(grepl("ʀ|ʁ", Phoneme))
``` 

```{r}
uvulars_with_rhotics %>%
  select(InventoryID, Glottocode, LanguageName, Phoneme, Allophones) %>%
  head() %>%
  kable()
```

```{r}
uvulars_with_rhotics %>%
  group_by(Phoneme) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  kable()
```

Test that they've all been removed.

```{r}
expect_equal(nrow(uvulars_no_rhotics %>% filter(grepl("ʀ|ʁ", Phoneme))), 0)
expect_equal(nrow(uvulars %>% filter(!grepl("ʀ|ʁ", Phoneme))) + nrow(uvulars %>% filter(grepl("ʀ|ʁ", Phoneme))), nrow(uvulars))
```

Count the uvulars and marginal uvulars without rhotics.

```{r}
uvular_counts_minus_rhotics <- uvulars_no_rhotics %>%
  group_by(InventoryID) %>%
  summarize(Uvulars_no_rhotics = n())

uvular_marginals_minus_rhotics <- uvulars_no_rhotics %>%
  filter(Marginal) %>%
  group_by(InventoryID) %>%
  summarize(Marginal_Uvular_no_rhotics = n())
```

Since we do some hand annotation, e.g. adding in Urban's more fine-grained macroareas, we need to load up the data that we use in our analyses and then add in the new non-rhotic uvular counts.

```{r}
results <- read_csv("../Data/uvulars_ejectives_pruned2.csv")
```

A quick look:

```{r}
results %>%
  head() %>%
  kable()
```

Now let's merge in the new counts.

```{r}
results_uvulars <- left_join(results, uvular_counts_minus_rhotics)
results_uvulars <- left_join(results_uvulars, uvular_marginals_minus_rhotics)
```

Have a look.

```{r}
results_uvulars %>%
  head() %>%
  kable()
```

Let's replace the NA with zeros.

```{r}
results_uvulars <- results_uvulars %>% mutate(Uvulars_no_rhotics = replace_na(Uvulars_no_rhotics, 0))
results_uvulars <- results_uvulars %>% mutate(Marginal_Uvular_no_rhotics = replace_na(Marginal_Uvular_no_rhotics, 0))
```

Let's check where the counts differ and make sure we're getting the same languages that we noted above with rhotic uvulars. There are 11 observations.

```{r}
results_uvulars %>%
  filter(Uvulars != Uvulars_no_rhotics) %>%
  select(InventoryID, Glottocode, GlottologName, SpecificDialect, Source, Uvulars, Uvulars_no_rhotics) %>%
  kable()
```

Write the non-rhotic uvular counts to disk.

```{r}
write_csv(results_uvulars, "../Data/uvulars_ejectives_pruned2_rhotics.csv")
```


# References
